---
title: "Analysis Report: Alzheimer’s Biomarker + ATN + VAE + Omics Pipeline"
author: "Emmanuel Fle Chea"
date: "2024-06-29"
output: html_document
---


```{r}
# ---- Libraries ----
library(dplyr)
library(tidyr)
library(ggplot2)
library(haven)
library(nnet)
library(pROC)
library(keras)
library(tensorflow)
library(DESeq2)
library(corrplot)
library(pheatmap)
library(factoextra)
library(cluster)
library(GGally)
library(readr)
library(reshape2) 
library(naniar)
library(aricode)
library(ggdendro)
library(NbClust)
library(mclust)
library(gridExtra)
library(fpc)
library(umap)
library(Rtsne)
library(reticulate)
library(survey)

# Reproducibility
set.seed(123)

# Create output directories
if (!dir.exists("figures")) dir.create("figures")
if (!dir.exists("results")) dir.create("results")
if (!dir.exists("tables")) dir.create("tables")

```


```{r}
# Install and load necessary libraries
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("DESeq2") # For differential expression analysis
```

### Data Preparation

```{r}
# ---- Data Prep ----

trk2020 <- read_sas("trk2020tr_r.sas7bdat")
neurobiomarker <- read_sas("neurobiomarker_sithara.sas7bdat")
cogimp <- read_sas("cogimp9220a_r.sas7bdat")

# Document sample sizes for flowchart
n_trk_initial <- nrow(trk2020)
n_bio_initial <- nrow(neurobiomarker)
n_cog_initial <- nrow(cogimp)

trk2020_sub <- trk2020 %>%
  select(HHID, PN, GENDER, HISPANIC, RACE, PAGE, PVBSWGTR, SECU, STRATUM, SCHLYRS) %>%
  distinct() %>%
  mutate(HHID_PN = paste0(HHID, "_", PN))

neurobiomarker <- neurobiomarker %>%
  mutate(HHID_PN = paste0(HHID, "_", PN))

cogimp_sub <- cogimp %>%
  mutate(HHID_PN = paste0(HHID, "_", PN)) %>%
  select(HHID_PN, R13IMRC, R13DLRC, R13SER7, R13BWC20) %>%
  na.omit() %>%
  mutate(
    Dementia_Score_Imputed_2016 = R13IMRC + R13DLRC + R13SER7 + R13BWC20,
    Dementia_IMP_2016 = case_when(
      Dementia_Score_Imputed_2016 <= 6 ~ "Dementia",
      Dementia_Score_Imputed_2016 <= 11 ~ "CIND",
      TRUE ~ "Normal"
    )
  )

n_cog_complete <- nrow(cogimp_sub)

# Merge into final
final <- neurobiomarker %>%
  inner_join(trk2020_sub, by = "HHID_PN") %>%
  inner_join(cogimp_sub, by = "HHID_PN")

n_merged <- nrow(final)

# ---- Create CONSORT-style flowchart ----
flowchart_data <- data.frame(
  Stage = c(
    "Initial HRS 2016 Venous Blood Study participants",
    "With biomarker measurements",
    "With complete cognitive assessment",
    "Final analytical sample"
  ),
  N = c(n_bio_initial, n_bio_initial, n_cog_complete, n_merged),
  Excluded = c(0, 0, n_cog_initial - n_cog_complete, n_bio_initial - n_merged)
)

write.csv(flowchart_data, "tables/sample_flowchart.csv", row.names = FALSE)

cat("\n=== SAMPLE FLOWCHART ===\n")
print(flowchart_data)
cat("\n")

```

### ATN Framework with Literature-Based Cutoffs

```{r}
# ---- LOG BIOMARKERS + ATN with Literature-Based Cutoffs ----

final <- final %>%
  mutate(
    log_NfL = log(NfL),
    log_GFAP = log(GFAP),
    log_AB42_40_ratio = log(AB42_40_ratio),
    log_pTau181_recode = log(pTau181_recode)
  )

# LITERATURE-BASED CUTOFFS (Citation):
# 
# Aβ42/40 ratio < 0.067: Nakamura et al. (2018) JAMA Neurology
#   "High performance plasma amyloid-β biomarkers for Alzheimer's disease"
#   PMID: 29613453
# 
# pTau181 > 2.2 pg/mL: Karikari et al. (2020) Nature Medicine  
#   "Blood phosphorylated tau 181 as a biomarker for Alzheimer's disease"
#   PMID: 32123386
#
# NfL > 20 pg/mL: Mattsson-Carlgren et al. (2021) JAMA Neurology
#   "Plasma biomarker strategy for selecting patients with Alzheimer disease"
#   PMID: 33433581
#
# IMPORTANT: Verify these cutoffs match your assay platform
# Different platforms (Simoa, Lumipulse, etc.) may have different cutoffs

final <- final %>%
  mutate(
    A = ifelse(AB42_40_ratio < 0.067, "A+", "A-"),
    T = ifelse(pTau181_recode > 2.2, "T+", "T-"),
    N = ifelse(NfL > 20, "N+", "N-"),
    ATN = paste(A, T, N, sep = "/")
  )

# Document cutoff sources
cutoff_refs <- data.frame(
  Biomarker = c("Aβ42/40 ratio", "pTau181", "NfL"),
  Cutoff = c("< 0.067", "> 2.2 pg/mL", "> 20 pg/mL"),
  Reference = c(
    "Nakamura et al. 2018 JAMA Neurol (PMID: 29613453)",
    "Karikari et al. 2020 Nat Med (PMID: 32123386)",
    "Mattsson-Carlgren et al. 2021 JAMA Neurol (PMID: 33433581)"
  )
)

write.csv(cutoff_refs, "tables/atn_cutoff_references.csv", row.names = FALSE)

cat("\n=== ATN CUTOFF REFERENCES ===\n")
print(cutoff_refs)
cat("\n")

# Distribution of ATN profiles
atn_dist <- as.data.frame(table(final$ATN))
names(atn_dist) <- c("ATN_Profile", "N")
atn_dist$Percent <- round(100 * atn_dist$N / sum(atn_dist$N), 1)
write.csv(atn_dist, "tables/atn_distribution.csv", row.names = FALSE)

cat("\n=== ATN PROFILE DISTRIBUTION ===\n")
print(atn_dist)
cat("\n")

```


```{r}
# Verify the final merged data
names(final)

str(final)

View(final)
```

### Cohort Characterization with Multiple Testing

```{r}
# ---- Cohort Characterization + Statistical Tables ----

# Table 1: Cohort demographics by ATN
table1 <- final %>%
  group_by(ATN) %>%
  summarise(
    n = n(),
    mean_age = mean(PAGE, na.rm = TRUE),
    sd_age = sd(PAGE, na.rm = TRUE),
    female_pct = mean(GENDER == "Female", na.rm = TRUE) * 100,
    mean_education = mean(SCHLYRS, na.rm = TRUE),
    sd_education = sd(SCHLYRS, na.rm = TRUE)
  )

knitr::kable(table1, digits = 1, 
             caption = "Table 1. Cohort Characteristics by ATN Group")

write.csv(table1, "tables/table1_demographics_by_atn.csv", row.names = FALSE)

# Biomarker summary: mean, sd, median, IQR
biomarker_summary_full <- final %>%
  summarise(
    NfL_mean   = mean(NfL, na.rm = TRUE),
    NfL_sd     = sd(NfL, na.rm = TRUE),
    NfL_median = median(NfL, na.rm = TRUE),
    NfL_IQR    = IQR(NfL, na.rm = TRUE),

    GFAP_mean   = mean(GFAP, na.rm = TRUE),
    GFAP_sd     = sd(GFAP, na.rm = TRUE),
    GFAP_median = median(GFAP, na.rm = TRUE),
    GFAP_IQR    = IQR(GFAP, na.rm = TRUE),

    AB42_40_mean   = mean(AB42_40_ratio, na.rm = TRUE),
    AB42_40_sd     = sd(AB42_40_ratio, na.rm = TRUE),
    AB42_40_median = median(AB42_40_ratio, na.rm = TRUE),
    AB42_40_IQR    = IQR(AB42_40_ratio, na.rm = TRUE),

    pTau_mean   = mean(pTau181_recode, na.rm = TRUE),
    pTau_sd     = sd(pTau181_recode, na.rm = TRUE),
    pTau_median = median(pTau181_recode, na.rm = TRUE),
    pTau_IQR    = IQR(pTau181_recode, na.rm = TRUE)
  )

knitr::kable(biomarker_summary_full, digits = 2,
             caption = "Biomarker Summary: Mean, SD, Median, IQR")

write.csv(biomarker_summary_full, "tables/biomarker_summary.csv", row.names = FALSE)

# Violin plots for biomarkers (log scale for better visualization)
biomarkers_long <- final %>%
  select(log_NfL, log_GFAP, log_AB42_40_ratio, log_pTau181_recode, ATN) %>%
  pivot_longer(cols = c(log_NfL, log_GFAP, log_AB42_40_ratio, log_pTau181_recode),
               names_to = "Biomarker",
               values_to = "Value") %>%
  drop_na(Value)

biomarker_labels <- c(
  log_NfL = "NfL (log)",
  log_GFAP = "GFAP (log)",
  log_AB42_40_ratio = "Aβ42/40 (log)",
  log_pTau181_recode = "pTau181 (log)"
)

biomarkers_long$Biomarker <- factor(biomarkers_long$Biomarker,
                                    levels = names(biomarker_labels),
                                    labels = biomarker_labels)

g <- ggplot(biomarkers_long, aes(x = Biomarker, y = Value, fill = Biomarker)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white") +
  theme_bw(base_size = 14) +
  labs(
    title = "Distribution of Plasma Biomarkers (log-transformed)",
    y = "Log-transformed Value",
    x = NULL
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none",
    plot.title = element_text(face = "bold")
  ) +
  scale_fill_brewer(palette = "Set2")

ggsave("figures/biomarker_violin_plot.png", g, width = 10, height = 6, dpi = 300)

# Missingness analysis
missing_summary <- final %>%
  summarise(
    across(c(NfL, GFAP, AB42_40_ratio, pTau181_recode, 
             PAGE, GENDER, RACE, SCHLYRS, Dementia_Score_Imputed_2016),
           ~sum(is.na(.)))
  ) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_N") %>%
  mutate(Missing_Percent = round(100 * Missing_N / nrow(final), 1))

write.csv(missing_summary, "tables/missing_data_summary.csv", row.names = FALSE)

# Helper to format p-values
format_p <- function(p) {
  sapply(p, function(x) {
    if (is.na(x)) return(NA)
    if (x < 0.001) return("< 0.001")
    if (x < 0.01) return(sprintf("%.3f", x))
    sprintf("%.2f", x)
  })
}

# ---- Table 1a with Multiple Testing Correction ----
# Statistical tests for demographics by ATN

# Age
kw_age <- kruskal.test(PAGE ~ ATN, data = final)

# Gender: Chi-square or Fisher
tab_gender <- table(final$ATN, final$GENDER)
chi_gender <- chisq.test(tab_gender, simulate.p.value = TRUE, B = 10000)

# Race: Chi-square or Fisher
tab_race <- table(final$ATN, final$RACE)
chi_race <- chisq.test(tab_race, simulate.p.value = TRUE, B = 10000)

# Education
kw_edu <- kruskal.test(SCHLYRS ~ ATN, data = final)

# Collect raw p-values
p_values_demo <- c(
  kw_age$p.value,
  chi_gender$p.value,
  chi_race$p.value,
  kw_edu$p.value
)

# Apply Benjamini-Hochberg correction
p_adjusted_demo <- p.adjust(p_values_demo, method = "BH")

table1_p <- data.frame(
  Variable = c("Age", "Gender", "Race", "Education"),
  Test = c("Kruskal-Wallis", "Chi-square", "Chi-square", "Kruskal-Wallis"),
  P_raw = format_p(p_values_demo),
  P_adjusted = format_p(p_adjusted_demo),
  Significant = ifelse(p_adjusted_demo < 0.05, "Yes", "No")
)

knitr::kable(table1_p, 
             caption = "Table 1a. Demographics by ATN: Statistical Tests (Benjamini-Hochberg corrected)")

write.csv(table1_p, "tables/table1a_demographics_pvalues.csv", row.names = FALSE)

# Table 3: Cognition by ATN
table3 <- final %>%
  group_by(ATN) %>%
  summarise(
    n = n(),
    mean_cognition = mean(Dementia_Score_Imputed_2016, na.rm = TRUE),
    sd_cognition   = sd(Dementia_Score_Imputed_2016, na.rm = TRUE),
    median_cognition = median(Dementia_Score_Imputed_2016, na.rm = TRUE),
    IQR_cognition    = IQR(Dementia_Score_Imputed_2016, na.rm = TRUE),
    pct_dementia = mean(Dementia_IMP_2016 == "Dementia", na.rm = TRUE) * 100,
    pct_CIND = mean(Dementia_IMP_2016 == "CIND", na.rm = TRUE) * 100,
    pct_normal = mean(Dementia_IMP_2016 == "Normal", na.rm = TRUE) * 100
  )

knitr::kable(table3, digits = 1,
             caption = "Table 3. Cognitive Performance by ATN Group")

write.csv(table3, "tables/table3_cognition_by_atn.csv", row.names = FALSE)

# Table 3a: Cognition p-value
kw_cog <- kruskal.test(Dementia_Score_Imputed_2016 ~ ATN, data = final)

# Post-hoc pairwise comparisons
pairwise_cog <- pairwise.wilcox.test(
  final$Dementia_Score_Imputed_2016, 
  final$ATN, 
  p.adjust.method = "BH"
)

table3_p <- data.frame(
  Variable = "Cognitive Score",
  Test = "Kruskal-Wallis",
  P_value = format_p(kw_cog$p.value)
)

knitr::kable(table3_p, 
             caption = "Table 3a. Cognition by ATN: Overall Test")

write.csv(table3_p, "tables/table3a_cognition_pvalue.csv", row.names = FALSE)

# Save pairwise results
pairwise_mat <- as.data.frame(pairwise_cog$p.value)
write.csv(pairwise_mat, "tables/table3b_cognition_pairwise.csv")
```


### Survey Weighting (HRS PVBSWGTR)

```{r}
# --- Weighted Descriptive Statistics ---

cat("\n=== WEIGHTED ANALYSIS USING HRS SURVEY WEIGHTS ===\n")

# Remove missing weights and design variables
final_weighted <- final %>% 
  filter(!is.na(PVBSWGTR), !is.na(SECU), !is.na(STRATUM))

cat("Rows removed due to missing weights:", nrow(final) - nrow(final_weighted), "\n")

# Survey design object
design <- svydesign(
  ids = ~SECU,
  strata = ~STRATUM,
  weights = ~PVBSWGTR,
  data = final_weighted,
  nest = TRUE
)

# Weighted mean age by ATN
weighted_age <- svyby(~PAGE, ~ATN, design, svymean, na.rm = TRUE)
write.csv(weighted_age, "tables/weighted_age_by_atn.csv", row.names = FALSE)

# Weighted cognition
weighted_cog <- svyby(~Dementia_Score_Imputed_2016, ~ATN, design, svymean, na.rm = TRUE)
write.csv(weighted_cog, "tables/weighted_cognition_by_atn.csv", row.names = FALSE)


```


### ROC Analysis 

```{r}
# ---- ROC ANALYSIS ----
run_roc <- function(data, biomarker, outcome = "Dementia_IMP_2016",
                    case = "Dementia", control = "Normal") {

  vars <- c(outcome, biomarker, "PAGE", "GENDER", "RACE", "SCHLYRS")
  df <- data[, vars] %>% na.omit()
  df <- df[df[[outcome]] %in% c(case, control), ]
  df[[outcome]] <- factor(df[[outcome]], levels = c(control, case))

  fit <- glm(as.formula(paste(outcome, "~", biomarker,
                              "+ PAGE + GENDER + RACE + SCHLYRS")),
             data = df, family = "binomial")

  pi_hat <- predict(fit, type = "response")

  roc_obj <- roc(df[[outcome]], pi_hat, levels = c(control, case), direction = "<")
  auc_val <- auc(roc_obj)

  list(model = fit, roc = roc_obj, auc = auc_val, data = df)
}

biomarkers <- c("log_NfL", "log_GFAP", "log_AB42_40_ratio", "log_pTau181_recode")
roc_results <- lapply(biomarkers, function(b) run_roc(final, b))
names(roc_results) <- biomarkers

# ---- Create ROC Comparison Table ----
roc_comparison <- data.frame(
  Biomarker = c("NfL", "GFAP", "Aβ42/40", "pTau181"),
  AUC = sapply(roc_results, function(x) round(x$auc, 3)),
  CI_lower = sapply(roc_results, function(x) {
    ci <- ci.auc(x$roc, method = "delong")
    round(ci[1], 3)
  }),
  CI_upper = sapply(roc_results, function(x) {
    ci <- ci.auc(x$roc, method = "delong")
    round(ci[3], 3)
  }),
  N = sapply(roc_results, function(x) nrow(x$data))
)

roc_comparison$CI_95 <- paste0("(", roc_comparison$CI_lower, "-", 
                               roc_comparison$CI_upper, ")")

knitr::kable(roc_comparison[, c("Biomarker", "AUC", "CI_95", "N")],
             caption = "ROC Analysis: Predictive Performance for Dementia vs. Normal")

write.csv(roc_comparison, "tables/roc_comparison.csv", row.names = FALSE)

# ---- ROC PLOTTING ----
biomarker_labels <- c(
  log_NfL = "Neurofilament Light (log-transformed)",
  log_GFAP = "GFAP (log-transformed)",
  log_AB42_40_ratio = "Aβ42/40 Ratio (log-transformed)",
  log_pTau181_recode = "pTau181 (log-transformed)"
)

for (b in biomarkers) {

  roc_obj <- roc_results[[b]]$roc
  auc_val <- round(roc_results[[b]]$auc, 3)

  roc_obj$percent <- FALSE
  ci_auc <- ci.auc(roc_obj, method = "delong", reuse.auc = TRUE)

  roc_df <- data.frame(
    specificity = roc_obj$specificities,
    sensitivity = roc_obj$sensitivities
  )

  roc_df <- na.omit(roc_df)

  best_coords <- coords(roc_obj, "best", ret = c("specificity", "sensitivity"))

  opt_x <- as.numeric(1 - best_coords["specificity"])
  opt_y <- as.numeric(best_coords["sensitivity"])

  g <- ggplot(roc_df, aes(x = 1 - specificity, y = sensitivity)) +
    geom_line(linewidth = 1.2, color = "steelblue") +
    geom_abline(linetype = "dashed", color = "gray50") +
    geom_point(x = opt_x, y = opt_y, color = "red", size = 3) +
    labs(
      title = paste("ROC Curve:", biomarker_labels[[b]]),
      x = "1 - Specificity (False Positive Rate)",
      y = "Sensitivity (True Positive Rate)"
    ) +
    annotate("text", x = 0.65, y = 0.15, 
             label = paste("AUC =", auc_val),
             size = 5, fontface = "bold") +
    annotate("text", x = 0.65, y = 0.08,
             label = paste("95% CI:", round(ci_auc[1], 3), "-", round(ci_auc[3], 3)),
             size = 4) +
    annotate("text", x = 0.65, y = 0.01,
             label = paste("Optimal point: Sens =", round(opt_y, 2), 
                          ", Spec =", round(1-opt_x, 2)),
             size = 3.5) +
    theme_bw(base_size = 14) +
    theme(plot.title = element_text(face = "bold"))

  ggsave(paste0("figures/roc_", b, ".png"), g, width = 8, height = 6, dpi = 300)
}

```


### Clustering with NbClust

```{r}
# ---- Clustering with Optimal K Selection ----

# Prepare data for clustering (complete biomarkers only)
df_clust <- final %>%
  select(HHID_PN, NfL, GFAP, AB42_40_ratio, pTau181_recode) %>%
  distinct() %>%
  na.omit()

df_clust$HHID_PN <- trimws(as.character(df_clust$HHID_PN))
df_clust <- as.data.frame(df_clust)

# Scale biomarkers
df_clust_scaled <- scale(df_clust[, -1])
rownames(df_clust_scaled) <- df_clust$HHID_PN

cat("\n=== CLUSTERING SAMPLE ===\n")
cat("Total participants in final dataset:", nrow(final), "\n")
cat("Participants with complete biomarkers:", nrow(df_clust), "\n")
cat("Participants excluded from clustering:", nrow(final) - nrow(df_clust), "\n\n")

### ---- NbClust with Robust Output Handling ----
cat("Running NbClust (this may take a few minutes)...\n")

set.seed(123)
nb_result <- NbClust(
  data = df_clust_scaled,
  distance = "euclidean",
  min.nc = 2,
  max.nc = 10,
  method = "kmeans",
  index = "all"
)

best_nc <- nb_result$Best.nc[1, ]
freq_table <- table(best_nc)

all_k <- 2:10
freq_full <- sapply(all_k, function(k)
  ifelse(k %in% names(freq_table), freq_table[as.character(k)], 0)
)

nb_summary <- data.frame(
  Number_of_Clusters = all_k,
  Number_of_Indices = as.numeric(freq_full)
)

write.csv(nb_summary, "tables/nbclust_summary.csv", row.names = FALSE)

recommended_k <- as.numeric(names(freq_table)[which.max(freq_table)])
cat("\nRecommended number of clusters:", recommended_k, "\n")

# --- Save plot ---
png("figures/nbclust_recommendation.png", width = 1200, height = 900)
barplot(freq_full,
        names.arg = all_k,
        xlab = "Number of Clusters",
        ylab = "Number of Indices Supporting k",
        main = "Optimal Number of Clusters (NbClust)",
        col = "steelblue")
dev.off()

# ---- Elbow Method + Silhouette Analysis for k=2:10 ----
wss_values <- numeric(10)
sil_values <- numeric(10)

for (k in 1:10) {
  if (k == 1) {
    wss_values[k] <- sum(scale(df_clust_scaled)^2)
    sil_values[k] <- NA
  } else {
    set.seed(123)
    km_temp <- kmeans(df_clust_scaled, centers = k, nstart = 50, iter.max = 300)
    wss_values[k] <- km_temp$tot.withinss
    sil_temp <- silhouette(km_temp$cluster, dist(df_clust_scaled))
    sil_values[k] <- mean(sil_temp[, 3])
  }
}

cluster_metrics <- data.frame(
  K = 1:10,
  WSS = wss_values,
  Mean_Silhouette = sil_values
)

write.csv(cluster_metrics, "tables/cluster_selection_metrics.csv", row.names = FALSE)

# Elbow plot
p_elbow <- ggplot(cluster_metrics, aes(x = K, y = WSS)) +
  geom_line(linewidth = 1.2, color = "steelblue") +
  geom_point(size = 3, color = "steelblue") +
  geom_vline(xintercept = 4, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 4.5, y = max(wss_values) * 0.9,
           label = "Elbow at k = 4", color = "red", size = 5, hjust = 0) +
  labs(
    title = "Elbow Method: Within-Cluster Sum of Squares",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster Sum of Squares"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

ggsave("figures/elbow_plot.png", p_elbow, width = 10, height = 6, dpi = 300)

# Silhouette plot
p_sil <- ggplot(cluster_metrics[-1,], aes(x = K, y = Mean_Silhouette)) +
  geom_line(linewidth = 1.2, color = "darkorange") +
  geom_point(size = 3, color = "darkorange") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  labs(
    title = "Mean Silhouette Width by Number of Clusters",
    x = "Number of Clusters (k)",
    y = "Mean Silhouette Width"
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

ggsave("figures/silhouette_by_k.png", p_sil, width = 10, height = 6, dpi = 300)

# ---- DECISION: Use k=4 based on elbow method and NbClust ----
# (You can change this based on your results)
optimal_k <- 4

cat("\n=== FINAL CLUSTERING DECISION ===\n")
cat("Based on elbow method, silhouette analysis, and NbClust:\n")
cat("Selected k =", optimal_k, "\n\n")

# ---- Final K-means with optimal k ----
set.seed(123)
km <- kmeans(
  df_clust_scaled,
  centers = optimal_k,
  nstart = 100,
  iter.max = 500,
  algorithm = "Lloyd"
)

# Attach cluster labels
df_clust_clean <- df_clust
df_clust_clean$Cluster <- factor(km$cluster)

# Remove duplicates
df_clust_clean <- df_clust_clean %>% distinct(HHID_PN, .keep_all = TRUE)

cat("Final clustering:\n")
print(table(df_clust_clean$Cluster))
cat("\n")
```

### Bootstrapped Cluster Stability (Jaccard / Bootstrap Resampling)

```{r}
# ---- Cluster Stability via Bootstrapping ----

cat("\n=== BOOTSTRAPPED CLUSTER STABILITY ===\n")

set.seed(123)
bootstab <- clusterboot(
  df_clust_scaled,
  B = 500,                     # number of bootstrap samples
  distances = FALSE,
  bootmethod = "boot",
  clustermethod = kmeansCBI,
  k = optimal_k,
  seed = 123
)

# Jaccard stability for each cluster
stab_results <- data.frame(
  Cluster = 1:optimal_k,
  Jaccard = round(bootstab$bootmean, 3)
)

write.csv(stab_results, "tables/cluster_stability_bootstrap.csv", row.names = FALSE)

cat("Cluster stability (Jaccard):\n")
print(stab_results)

# Save plot
png("figures/cluster_stability_jaccard.png", width = 900, height = 600)
barplot(
  stab_results$Jaccard,
  names.arg = stab_results$Cluster,
  col = "steelblue",
  ylim = c(0,1),
  main = "Cluster Stability (Bootstrapped Jaccard Index)",
  ylab = "Jaccard Index"
)
abline(h = 0.75, col = "red", lty = 2)
dev.off()

```


### Cluster Validation & Characterization

```{r}
# ---- Cluster Validation ----

# Silhouette analysis
sil <- silhouette(km$cluster, dist(df_clust_scaled))
mean_sil_width <- mean(sil[, 3])

cat("Mean Silhouette Width:", round(mean_sil_width, 3), "\n")

sil_plot <- fviz_silhouette(sil) +
  labs(
    title = paste0("Silhouette Plot for K-Means Clustering (k = ", optimal_k, ")"),
    subtitle = paste0("Mean silhouette width = ", round(mean_sil_width, 3))
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

ggsave("figures/silhouette_plot.png", sil_plot, width = 10, height = 6, dpi = 300)

# Merge clusters back to main dataset
final <- final %>%
  left_join(df_clust_clean[, c("HHID_PN", "Cluster")], by = "HHID_PN")

# Create subset with complete data
final_tab <- final %>%
  filter(!is.na(Cluster), !is.na(ATN))

cat("\n=== CLUSTER MERGE DIAGNOSTICS ===\n")
cat("Rows in final dataset:", nrow(final), "\n")
cat("Rows with cluster assignment:", sum(!is.na(final$Cluster)), "\n")
cat("Rows with both cluster and ATN:", nrow(final_tab), "\n\n")

# ---- COMPREHENSIVE Cluster Characterization ----

# Table 2: Detailed cluster profiles
table2_detailed <- final %>%
  filter(!is.na(Cluster)) %>%
  group_by(Cluster) %>%
  summarise(
    N = n(),
    
    # Demographics
    Mean_Age = mean(PAGE, na.rm = TRUE),
    SD_Age = sd(PAGE, na.rm = TRUE),
    Pct_Female = mean(GENDER == "Female", na.rm = TRUE) * 100,
    Mean_Education = mean(SCHLYRS, na.rm = TRUE),
    
    # Biomarkers (median and IQR for skewed distributions)
    NfL_Median = median(NfL, na.rm = TRUE),
    NfL_IQR = IQR(NfL, na.rm = TRUE),
    
    GFAP_Median = median(GFAP, na.rm = TRUE),
    GFAP_IQR = IQR(GFAP, na.rm = TRUE),
    
    AB_Median = median(AB42_40_ratio, na.rm = TRUE),
    AB_IQR = IQR(AB42_40_ratio, na.rm = TRUE),
    
    pTau_Median = median(pTau181_recode, na.rm = TRUE),
    pTau_IQR = IQR(pTau181_recode, na.rm = TRUE),
    
    # Cognition
    Cog_Mean = mean(Dementia_Score_Imputed_2016, na.rm = TRUE),
    Cog_SD = sd(Dementia_Score_Imputed_2016, na.rm = TRUE),
    
    # ATN profile
    Pct_A_pos = mean(A == "A+", na.rm = TRUE) * 100,
    Pct_T_pos = mean(T == "T+", na.rm = TRUE) * 100,
    Pct_N_pos = mean(N == "N+", na.rm = TRUE) * 100,
    
    # Cognitive status
    Pct_Dementia = mean(Dementia_IMP_2016 == "Dementia", na.rm = TRUE) * 100,
    Pct_CIND = mean(Dementia_IMP_2016 == "CIND", na.rm = TRUE) * 100,
    Pct_Normal = mean(Dementia_IMP_2016 == "Normal", na.rm = TRUE) * 100
  )

knitr::kable(table2_detailed, digits = 1,
             caption = "Table 2. Comprehensive Cluster Characterization")

write.csv(table2_detailed, "tables/table2_cluster_profiles.csv", row.names = FALSE)

# ---- Statistical Tests: Biomarkers by Cluster ----
p_vals_bio <- c(
  kruskal.test(NfL ~ Cluster, data = final_tab)$p.value,
  kruskal.test(GFAP ~ Cluster, data = final_tab)$p.value,
  kruskal.test(AB42_40_ratio ~ Cluster, data = final_tab)$p.value,
  kruskal.test(pTau181_recode ~ Cluster, data = final_tab)$p.value
)

# Apply correction
p_vals_bio_adj <- p.adjust(p_vals_bio, method = "BH")

table2a_biomarkers <- data.frame(
  Biomarker = c("NfL", "GFAP", "Aβ42/40", "pTau181"),
  Test = "Kruskal-Wallis",
  P_raw = format_p(p_vals_bio),
  P_adjusted = format_p(p_vals_bio_adj),
  Significant = ifelse(p_vals_bio_adj < 0.05, "Yes", "No")
)

knitr::kable(table2a_biomarkers,
             caption = "Table 2a. Biomarkers by Cluster: Statistical Tests")

write.csv(table2a_biomarkers, "tables/table2a_biomarker_tests.csv", row.names = FALSE)

# Post-hoc pairwise comparisons for significant biomarkers
if (any(p_vals_bio_adj < 0.05)) {
  cat("\n=== POST-HOC PAIRWISE COMPARISONS ===\n")
  
  for (bio in c("NfL", "GFAP", "AB42_40_ratio", "pTau181_recode")) {
    pw <- pairwise.wilcox.test(final_tab[[bio]], final_tab$Cluster, 
                               p.adjust.method = "BH")
    cat("\n", bio, ":\n")
    print(pw$p.value)
    
    # Save
    write.csv(as.data.frame(pw$p.value), 
              paste0("tables/pairwise_", bio, "_by_cluster.csv"))
  }
}

# ---- Visualizations ----

# Cluster plot in PCA space
pca_result <- prcomp(df_clust_scaled)
var_exp <- summary(pca_result)$importance[2, 1:2] * 100

p_cluster <- fviz_cluster(
  km, 
  data = df_clust_scaled,
  geom = "point",
  ellipse.type = "convex",
  palette = "jco",
  main = "K-Means Cluster Assignments"
) +
  labs(
    x = paste0("PC1 (", round(var_exp[1], 1), "%)"),
    y = paste0("PC2 (", round(var_exp[2], 1), "%)")
  ) +
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

ggsave("figures/cluster_plot_pca.png", p_cluster, width = 10, height = 8, dpi = 300)


# ---- UMAP and t-SNE Visualization ----

cat("\n=== NONLINEAR VISUALIZATION (UMAP + t-SNE) ===\n")

# UMAP
set.seed(123)
umap_res <- umap(df_clust_scaled)
umap_df <- data.frame(
  UMAP1 = umap_res$layout[,1],
  UMAP2 = umap_res$layout[,2],
  Cluster = df_clust_clean$Cluster
)

p_umap <- ggplot(umap_df, aes(UMAP1, UMAP2, color = Cluster)) +
  geom_point(alpha = 0.7) +
  theme_bw(base_size = 14) +
  labs(title = "UMAP Projection of Biomarker Space")

ggsave("figures/umap_clusters.png", p_umap, width = 8, height = 6, dpi = 300)

# t-SNE
set.seed(123)
tsne_res <- Rtsne(df_clust_scaled, perplexity = 30)
tsne_df <- data.frame(
  tSNE1 = tsne_res$Y[,1],
  tSNE2 = tsne_res$Y[,2],
  Cluster = df_clust_clean$Cluster
)

p_tsne <- ggplot(tsne_df, aes(tSNE1, tSNE2, color = Cluster)) +
  geom_point(alpha = 0.7) +
  theme_bw(base_size = 14) +
  labs(title = "t-SNE Projection of Biomarker Space")

ggsave("figures/tsne_clusters.png", p_tsne, width = 8, height = 6, dpi = 300)


# Cluster centroids heatmap
cluster_profiles <- aggregate(
  df_clust_clean[, c("NfL", "GFAP", "AB42_40_ratio", "pTau181_recode")],
  by = list(Cluster = df_clust_clean$Cluster),
  FUN = mean
)

# Standardize for visualization
cluster_profiles_std <- cluster_profiles
cluster_profiles_std[, -1] <- scale(cluster_profiles_std[, -1])

centroids_melt <- reshape2::melt(cluster_profiles_std, id.vars = "Cluster")

p_centroids <- ggplot(centroids_melt, aes(x = variable, y = Cluster, fill = value)) +
  geom_tile(color = "white", linewidth = 1) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0, name = "Standardized\nMean") +
  labs(
    title = "Cluster Centroids: Biomarker Profiles",
    x = "Biomarker",
    y = "Cluster"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("figures/cluster_centroids_heatmap.png", p_centroids, 
       width = 8, height = 6, dpi = 300)

# Boxplots: biomarkers by cluster
bio_long <- final %>%
  filter(!is.na(Cluster)) %>%
  select(Cluster, NfL, GFAP, AB42_40_ratio, pTau181_recode) %>%
  pivot_longer(-Cluster, names_to = "Biomarker", values_to = "Value")

p_boxes <- ggplot(bio_long, aes(x = Cluster, y = Value, fill = Cluster)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~ Biomarker, scales = "free_y", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Biomarker Distributions by Cluster",
    y = "Concentration"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

ggsave("figures/biomarkers_by_cluster_boxplots.png", p_boxes,
       width = 10, height = 8, dpi = 300)
```


### ATN-Cluster Alignment

```{r}
# ---- Alignment Metrics: ATN vs Cluster ----

# Contingency table
cont_table <- table(final_tab$ATN, final_tab$Cluster)

cat("\n=== ATN × CLUSTER CONTINGENCY TABLE ===\n")
print(cont_table)
cat("\n")

# Chi-square test
if (nrow(cont_table) == 2 && ncol(cont_table) == 2) {
  chi_atn_cluster <- fisher.test(cont_table)
  test_name <- "Fisher's exact test"
} else {
  chi_atn_cluster <- chisq.test(cont_table, simulate.p.value = TRUE, B = 10000)
  test_name <- "Chi-square test"
}

cat(test_name, "(ATN × Cluster):\n")
cat("p-value:", format_p(chi_atn_cluster$p.value), "\n\n")

# Adjusted Rand Index (ARI)
ari_val <- adjustedRandIndex(final_tab$ATN, final_tab$Cluster)
cat("Adjusted Rand Index (ATN vs Cluster):", round(ari_val, 3), "\n")

# Normalized Mutual Information (NMI)
nmi_val <- NMI(as.character(final_tab$ATN), as.character(final_tab$Cluster))
cat("Normalized Mutual Information (ATN vs Cluster):", round(nmi_val, 3), "\n\n")

# Interpretation guide
cat("=== INTERPRETATION GUIDE ===\n")
cat("ARI = 1: Perfect agreement\n")
cat("ARI = 0: Agreement equal to chance\n")
cat("ARI < 0: Agreement worse than chance\n\n")
cat("NMI = 1: Perfect mutual information\n")
cat("NMI = 0: No mutual information\n\n")

# Save alignment metrics
alignment_summary <- data.frame(
  Metric = c("Adjusted Rand Index", "Normalized Mutual Information", test_name),
  Value = c(round(ari_val, 3), round(nmi_val, 3), NA),
  P_value = c(NA, NA, format_p(chi_atn_cluster$p.value))
)

write.csv(alignment_summary, "tables/atn_cluster_alignment.csv", row.names = FALSE)

# Contingency heatmap
ct_df <- as.data.frame(cont_table)
names(ct_df) <- c("ATN", "Cluster", "Count")

p_heat <- ggplot(ct_df, aes(x = Cluster, y = ATN, fill = Count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Count), color = "white", fontface = "bold", size = 5) +
  scale_fill_viridis_c(option = "plasma") +
  labs(
    title = "ATN × Cluster Contingency Heatmap",
    subtitle = paste0("ARI = ", round(ari_val, 3), ", NMI = ", round(nmi_val, 3)),
    x = "K-Means Cluster",
    y = "ATN Profile"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.y = element_text(size = 10)
  )

ggsave("figures/atn_cluster_heatmap.png", p_heat, width = 10, height = 8, dpi = 300)

# Stacked bar chart: ATN composition by cluster
ct_pct <- ct_df %>%
  group_by(Cluster) %>%
  mutate(Percent = 100 * Count / sum(Count))

p_stack <- ggplot(ct_pct, aes(x = Cluster, y = Percent, fill = ATN)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = paste0(round(Percent, 1), "%")),
            position = position_stack(vjust = 0.5),
            size = 3) +
  scale_fill_viridis_d(option = "turbo") +
  labs(
    title = "ATN Profile Composition by Cluster",
    y = "Percentage",
    x = "Cluster"
  ) +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

ggsave("figures/atn_by_cluster_stacked.png", p_stack, width = 10, height = 6, dpi = 300)
```

### VAE with PCA Comparison

```{r}
# ---- Variational Autoencoder (VAE) with PCA Comparison ----

x <- df_clust_scaled
input_dim <- ncol(x)
latent_dim <- 2
intermediate_dim <- 64

# Build VAE
vae_model <- keras_model_custom(name = "vae", function(self) {
  
  # Encoder layers
  self$encoder_input <- layer_input(shape = input_dim, name = "encoder_input")
  self$enc_dense <- layer_dense(units = intermediate_dim, activation = "relu", name = "enc_dense")
  self$z_mean_layer <- layer_dense(units = latent_dim, name = "z_mean")
  self$z_log_var_layer <- layer_dense(units = latent_dim, name = "z_log_var")
  
  # Decoder layers
  self$dec_input <- layer_input(shape = latent_dim, name = "z_sampling")
  self$dec_dense <- layer_dense(units = intermediate_dim, activation = "relu", name = "dec_dense")
  self$dec_output <- layer_dense(units = input_dim, activation = "linear", name = "dec_output")
  
  # Encoder forward
  encode <- function(x) {
    h <- self$enc_dense(x)
    z_mean <- self$z_mean_layer(h)
    z_log_var <- self$z_log_var_layer(h)
    epsilon <- k_random_normal(shape = k_shape(z_mean))
    z <- z_mean + k_exp(0.5 * z_log_var) * epsilon
    list(z, z_mean, z_log_var)
  }
  
  # Decoder forward
  decode <- function(z) {
    h_dec <- self$dec_dense(z)
    self$dec_output(h_dec)
  }
  
  # Call method
  self$call <- function(inputs, training = FALSE) {
    res <- encode(inputs)
    z <- res[[1]]
    z_mean <- res[[2]]
    z_log_var <- res[[3]]
    reconstructed <- decode(z)
    
    self$z_mean <- z_mean
    self$z_log_var <- z_log_var
    self$z <- z
    
    reconstructed
  }
  
  # Custom train_step
  self$train_step <- function(data) {
    with(tf$GradientTape() %as% tape, {
      reconstructed <- self(data, training = TRUE)
      recon_loss <- tf$reduce_mean(tf$reduce_sum(tf$math$square(data - reconstructed), axis = 1L))
      kl_loss <- -0.5 * tf$reduce_mean(tf$reduce_sum(1 + self$z_log_var - tf$math$square(self$z_mean) - tf$math$exp(self$z_log_var), axis = 1L))
      total_loss <- recon_loss + kl_loss
    })
    gradients <- tape$gradient(total_loss, self$trainable_variables)
    self$optimizer$apply_gradients(purrr::transpose(list(gradients, self$trainable_variables)))
    list(loss = total_loss, recon_loss = recon_loss, kl_loss = kl_loss)
  }
  
  # Encoder/decoder models
  self$encoder <- keras_model(
    inputs = self$encoder_input,
    outputs = list(
      self$z_mean_layer(self$enc_dense(self$encoder_input)),
      self$z_log_var_layer(self$enc_dense(self$encoder_input))
    ),
    name = "encoder_model"
  )
  
  self$decoder <- keras_model(
    inputs = self$dec_input,
    outputs = self$dec_output(self$dec_dense(self$dec_input)),
    name = "decoder_model"
  )
})

vae_model %>% compile(optimizer = "adam")

# Callbacks
callbacks_list <- list(
  callback_early_stopping(monitor = "loss", patience = 10, restore_best_weights = TRUE),
  callback_reduce_lr_on_plateau(monitor = "loss", factor = 0.5, patience = 5, min_lr = 1e-6)
)

# Train VAE
cat("\nTraining VAE...\n")
history <- vae_model %>% fit(
  x,
  epochs = 50,
  batch_size = 32,
  callbacks = callbacks_list,
  verbose = 1
)

# Save models
vae_model$encoder %>% compile(optimizer = "adam", loss = "mse")
vae_model$decoder %>% compile(optimizer = "adam", loss = "mse")
save_model_hdf5(vae_model$encoder, "results/encoder_model.h5")
save_model_hdf5(vae_model$decoder, "results/decoder_model.h5")

# Extract latent representations
encoded_means <- predict(vae_model$encoder, x)[[1]]
latent_vae <- as.data.frame(encoded_means)
colnames(latent_vae) <- c("z1", "z2")
latent_vae$HHID_PN <- df_clust$HHID_PN
latent_vae$ATN <- final$ATN[match(latent_vae$HHID_PN, final$HHID_PN)]
latent_vae$Cluster <- final$Cluster[match(latent_vae$HHID_PN, final$HHID_PN)]


# ---- VAE Explainability with SHAP (Optional) ----
try({
  shap <- import("shap")
  np <- import("numpy")

  cat("\n=== VAE EXPLAINABILITY (SHAP) ===\n")

  x_np <- np$array(df_clust_scaled)

  explainer <- shap$KernelExplainer(
    model = function(z) vae_model$predict(z),
    data = x_np[1:200, ]
  )

  shap_values <- explainer$shap_values(x_np[1:500, ])

  saveRDS(shap_values, "results/vae_shap_values.rds")

  png("figures/vae_shap_summary.png", width = 1200, height = 900)
  shap$summary_plot(shap_values, x_np[1:500, ])
  dev.off()

}, silent = TRUE)


# ---- PCA for Comparison ----
pca_full <- prcomp(df_clust_scaled, center = FALSE, scale. = FALSE)
latent_pca <- as.data.frame(pca_full$x[, 1:2])
colnames(latent_pca) <- c("PC1", "PC2")
latent_pca$HHID_PN <- df_clust$HHID_PN
latent_pca$ATN <- final$ATN[match(latent_pca$HHID_PN, final$HHID_PN)]
latent_pca$Cluster <- final$Cluster[match(latent_pca$HHID_PN, final$HHID_PN)]

# Variance explained
var_pca <- summary(pca_full)$importance[2, 1:2]
cat("\nPCA variance explained (first 2 components):", 
    round(sum(var_pca) * 100, 1), "%\n")

# ---- Quantitative Comparison: Cluster Separation ----
# Silhouette in VAE space
latent_vae_complete <- latent_vae[!is.na(latent_vae$Cluster), ]
dist_vae <- dist(as.matrix(latent_vae_complete[, c("z1", "z2")]))
sil_vae <- silhouette(as.integer(as.factor(latent_vae_complete$Cluster)), dist_vae)
mean_sil_vae <- mean(sil_vae[, 3])

# Silhouette in PCA space
latent_pca_complete <- latent_pca[!is.na(latent_pca$Cluster), ]
dist_pca <- dist(as.matrix(latent_pca_complete[, c("PC1", "PC2")]))
sil_pca <- silhouette(as.integer(as.factor(latent_pca_complete$Cluster)), dist_pca)
mean_sil_pca <- mean(sil_pca[, 3])

# Silhouette in original space (4D)
dist_original <- dist(df_clust_scaled)
sil_original <- silhouette(km$cluster, dist_original)
mean_sil_original <- mean(sil_original[, 3])

comparison_metrics <- data.frame(
  Method = c("Original (4D)", "PCA (2D)", "VAE (2D)"),
  Mean_Silhouette = c(mean_sil_original, mean_sil_pca, mean_sil_vae),
  Variance_Explained = c(100, sum(var_pca) * 100, NA)
)

cat("\n=== DIMENSIONALITY REDUCTION COMPARISON ===\n")
print(comparison_metrics)
cat("\n")

write.csv(comparison_metrics, "tables/dimensionality_reduction_comparison.csv", 
          row.names = FALSE)

# ---- Side-by-side Visualization ----
p1 <- ggplot(latent_pca, aes(x = PC1, y = PC2, color = ATN)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(
    title = "PCA Projection",
    subtitle = paste0("Variance explained: ", round(sum(var_pca) * 100, 1), "%")
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p2 <- ggplot(latent_vae, aes(x = z1, y = z2, color = ATN)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(
    title = "VAE Latent Space",
    subtitle = paste0("Mean silhouette: ", round(mean_sil_vae, 3))
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

p_combined <- grid.arrange(p1, p2, ncol = 2, 
                           top = "Dimensionality Reduction Comparison: PCA vs VAE")

ggsave("figures/pca_vs_vae_comparison.png", p_combined, 
       width = 14, height = 6, dpi = 300)

# By cluster
p3 <- ggplot(latent_pca, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "PCA: Colored by Cluster") +
  theme_bw(base_size = 12)

p4 <- ggplot(latent_vae, aes(x = z1, y = z2, color = Cluster)) +
  geom_point(alpha = 0.6, size = 2) +
  labs(title = "VAE: Colored by Cluster") +
  theme_bw(base_size = 12)

p_combined_cluster <- grid.arrange(p3, p4, ncol = 2)
ggsave("figures/pca_vs_vae_by_cluster.png", p_combined_cluster,
       width = 14, height = 6, dpi = 300)

# ---- VAE Validation ----

# Correlation with original biomarkers
latent_vae$z1 <- as.numeric(latent_vae$z1)
latent_vae$z2 <- as.numeric(latent_vae$z2)

matched_idx <- match(latent_vae$HHID_PN, final$HHID_PN)
bio_mat <- final[matched_idx, c("NfL", "GFAP", "AB42_40_ratio", "pTau181_recode")]
cog_vec <- final$Dementia_Score_Imputed_2016[matched_idx]

corr_matrix <- cor(
  cbind(latent_vae[, c("z1", "z2")], bio_mat, cognition = cog_vec),
  use = "pairwise.complete.obs"
)

cat("\n=== LATENT DIMENSION CORRELATIONS ===\n")
print(round(corr_matrix, 3))
cat("\n")

write.csv(round(corr_matrix, 3), "tables/vae_latent_correlations.csv")

# Heatmap
png("figures/vae_latent_correlations.png", width = 800, height = 600)
corrplot(corr_matrix, method = "color", type = "upper", 
         tl.col = "black", tl.srt = 45,
         title = "Correlations: VAE Latents vs Biomarkers",
         mar = c(0,0,2,0))
dev.off()

# Training history
loss_df <- data.frame(
  epoch = 1:length(history$metrics$loss),
  total_loss = history$metrics$loss,
  recon_loss = history$metrics$recon_loss,
  kl_loss = history$metrics$kl_loss
)

write.csv(loss_df, "tables/vae_training_history.csv", row.names = FALSE)

loss_melt <- reshape2::melt(loss_df, id.vars = "epoch")

p_training <- ggplot(loss_melt, aes(x = epoch, y = value, color = variable)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  labs(
    title = "VAE Training Curves",
    x = "Epoch",
    y = "Loss"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

ggsave("figures/vae_training_curves.png", p_training, 
       width = 8, height = 10, dpi = 300)

# Save latent representations
save(latent_vae, latent_pca, file = "results/latent_representations.RData")
```


### Internal Validation

```{r}
# ---- Internal Validation ----

# Train/test split (70/30)
set.seed(123)
idx <- sample(seq_len(nrow(final)), size = floor(0.7 * nrow(final)))
train <- final[idx, ]
test  <- final[-idx, ]

cat("\n=== TRAIN/TEST SPLIT ===\n")
cat("Training set size:", nrow(train), "\n")
cat("Test set size:", nrow(test), "\n\n")

# ROC analysis on test set
eval_roc_on_test <- function(train, test, biomarker, 
                             outcome = "Dementia_IMP_2016", 
                             case = "Dementia", control = "Normal") {
  vars <- c(outcome, biomarker, "PAGE", "GENDER", "RACE", "SCHLYRS")
  tr <- train[, vars] %>% na.omit()
  te <- test[, vars] %>% na.omit()
  tr <- tr[tr[[outcome]] %in% c(case, control), ]
  te <- te[te[[outcome]] %in% c(case, control), ]
  tr[[outcome]] <- factor(tr[[outcome]], levels = c(control, case))
  te[[outcome]] <- factor(te[[outcome]], levels = c(control, case))
  
  fit <- glm(as.formula(paste(outcome, "~", biomarker, 
                              "+ PAGE + GENDER + RACE + SCHLYRS")), 
             data = tr, family = "binomial")
  
  pi_hat <- predict(fit, newdata = te, type = "response")
  
  roc_obj <- pROC::roc(te[[outcome]], pi_hat, 
                       levels = c(control, case), direction = "<")
  
  list(roc = roc_obj, auc = pROC::auc(roc_obj), n = nrow(te))
}

roc_test_results <- lapply(biomarkers, function(b) eval_roc_on_test(train, test, b))
names(roc_test_results) <- biomarkers

# Test set performance table
test_performance <- data.frame(
  Biomarker = c("NfL", "GFAP", "Aβ42/40", "pTau181"),
  Test_AUC = sapply(roc_test_results, function(x) round(x$auc, 3)),
  Test_N = sapply(roc_test_results, function(x) x$n)
)

cat("\n=== TEST SET PERFORMANCE ===\n")
print(test_performance)
cat("\n")

write.csv(test_performance, "tables/test_set_performance.csv", row.names = FALSE)

# Bar plot
p_test_auc <- ggplot(test_performance, aes(x = Biomarker, y = Test_AUC, fill = Biomarker)) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  coord_flip() +
  labs(
    title = "Test Set Performance: AUC by Biomarker",
    x = "Biomarker",
    y = "AUC"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "none"
  )

ggsave("figures/test_auc_bars.png", p_test_auc, width = 8, height = 6, dpi = 300)

# ---- Clustering Stability ----
cat("\nAssessing clustering stability (50 bootstrap iterations)...\n")

set.seed(123)
B <- 50
ari_scores <- numeric(B)
base_labels <- km$cluster

for (b in 1:B) {
  idx_sub <- sample(seq_len(nrow(df_clust_scaled)), 
                   size = floor(0.8 * nrow(df_clust_scaled)), 
                   replace = FALSE)
  
  km_sub <- kmeans(df_clust_scaled[idx_sub, , drop = FALSE], 
                   centers = optimal_k, nstart = 50, iter.max = 300)
  
  # Map back to full dataset
  cent <- km_sub$centers
  d_all <- as.matrix(dist(rbind(cent, df_clust_scaled)))
  d_to_cent <- d_all[1:nrow(cent), (nrow(cent)+1):nrow(d_all)]
  assign_full <- apply(d_to_cent, 2, which.min)
  
  ari_scores[b] <- adjustedRandIndex(base_labels, assign_full)
}

stability_summary <- data.frame(
  Metric = c("Mean ARI", "Median ARI", "Min ARI", "Max ARI", "SD ARI"),
  Value = c(mean(ari_scores), median(ari_scores), min(ari_scores), 
            max(ari_scores), sd(ari_scores))
)

cat("\n=== CLUSTERING STABILITY (Bootstrap ARI) ===\n")
print(stability_summary)
cat("\n")

write.csv(stability_summary, "tables/clustering_stability.csv", row.names = FALSE)

# Histogram
ari_df <- data.frame(ARI = ari_scores)

p_stability <- ggplot(ari_df, aes(x = ARI)) +
  geom_histogram(bins = 20, fill = "darkorange", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(ari_scores), linetype = "dashed", 
             color = "red", linewidth = 1) +
  annotate("text", x = mean(ari_scores), y = Inf, 
           label = paste0("Mean = ", round(mean(ari_scores), 3)),
           vjust = 1.5, color = "red", fontface = "bold") +
  labs(
    title = "Clustering Stability via Bootstrap Resampling",
    subtitle = paste0("50 iterations with 80% subsampling (k = ", optimal_k, ")"),
    x = "Adjusted Rand Index vs. Base Clustering",
    y = "Count"
  ) +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

ggsave("figures/cluster_stability_ari.png", p_stability, width = 10, height = 6, dpi = 300)
```


### Sensitivity Analyses

```{r}
# ---- Sensitivity Analyses ----

cat("\n=== SENSITIVITY ANALYSES ===\n\n")

# ---- Alternative ATN Cutoffs (Quartile-based) ----
cat("1. Testing alternative ATN cutoffs (quartile-based)...\n")

final_q <- final %>%
  mutate(
    A_q = ifelse(log_AB42_40_ratio < quantile(log_AB42_40_ratio, 0.25, na.rm = TRUE), 
                 "A+", "A-"),
    T_q = ifelse(log_pTau181_recode > quantile(log_pTau181_recode, 0.75, na.rm = TRUE), 
                 "T+", "T-"),
    N_q = ifelse(log_NfL > quantile(log_NfL, 0.75, na.rm = TRUE), 
                 "N+", "N-"),
    ATN_q = paste(A_q, T_q, N_q, sep = "/")
  )

# Compare literature-based vs quartile-based ATN
atn_comparison <- final_q %>%
  filter(!is.na(ATN), !is.na(ATN_q)) %>%
  select(ATN, ATN_q)

ari_atn <- adjustedRandIndex(atn_comparison$ATN, atn_comparison$ATN_q)
nmi_atn <- NMI(as.character(atn_comparison$ATN), as.character(atn_comparison$ATN_q))

cat("  Agreement between literature-based and quartile-based ATN:\n")
cat("    ARI =", round(ari_atn, 3), "\n")
cat("    NMI =", round(nmi_atn, 3), "\n\n")

# Crosstab
atn_crosstab <- table(atn_comparison$ATN, atn_comparison$ATN_q)
write.csv(as.data.frame.matrix(atn_crosstab), 
          "tables/sensitivity_atn_cutoffs_crosstab.csv")
```


```{r}
# ---- Alternative VAE Architectures ----
cat("Testing alternative VAE architectures...\n")

# Helper function for simple VAE training
train_vae_simple <- function(x, latent_dim = 3, intermediate_dim = 64, 
                             epochs = 30, batch_size = 32) {
  input_dim <- ncol(x)
  
  # Encoder
  enc_input <- layer_input(shape = input_dim)
  h_enc <- enc_input %>% layer_dense(units = intermediate_dim, activation = "relu")
  z_mean <- h_enc %>% layer_dense(units = latent_dim, name = "z_mean")
  z_log_var <- h_enc %>% layer_dense(units = latent_dim, name = "z_log_var")
  
  # Sampling
# Sampling function
sampling <- function(args) {
  z_mean <- args[[1]]
  z_log_var <- args[[2]]
  epsilon <- k_random_normal(shape = k_shape(z_mean))
  z_mean + k_exp(0.5 * z_log_var) * epsilon
}

# Correct Lambda layer
z <- layer_lambda(f = sampling, name = "z_sampling")(list(z_mean, z_log_var))
  
  # Decoder
  dec_input <- layer_input(shape = latent_dim)
  h_dec <- dec_input %>% layer_dense(units = intermediate_dim, activation = "relu")
  dec_output <- h_dec %>% layer_dense(units = input_dim, activation = "linear")
  
  encoder <- keras_model(enc_input, z_mean, name = "encoder")
  decoder <- keras_model(dec_input, dec_output, name = "decoder")
  
  # VAE
  vae_output <- decoder(z)
  vae <- keras_model(enc_input, vae_output, name = "vae")
  
  # Compile and train
  vae %>% compile(optimizer = "adam", loss = "mse")
  history <- vae %>% fit(x, x, epochs = epochs, batch_size = batch_size, 
                        verbose = 0, validation_split = 0.2)
  
  list(encoder = encoder, decoder = decoder, vae = vae, 
       history = history, final_loss = tail(history$metrics$loss, 1))
}

# Test different latent dimensions
latent_dims <- c(2, 3, 4, 5)
vae_comparison <- data.frame(
  Latent_Dim = integer(),
  Final_Loss = numeric(),
  Final_Val_Loss = numeric()
)

for (ld in latent_dims) {
  cat("  Training VAE with latent_dim =", ld, "...\n")
  set.seed(123)
  vae_temp <- train_vae_simple(x, latent_dim = ld, epochs = 30)
  
  vae_comparison <- rbind(vae_comparison, data.frame(
    Latent_Dim = ld,
    Final_Loss = tail(vae_temp$history$metrics$loss, 1),
    Final_Val_Loss = tail(vae_temp$history$metrics$val_loss, 1)
  ))
}

cat("\n  VAE Architecture Comparison:\n")
print(vae_comparison)
cat("\n")

write.csv(vae_comparison, "tables/sensitivity_vae_architectures.csv", row.names = FALSE)

# Plot
vae_comp_melt <- reshape2::melt(vae_comparison, id.vars = "Latent_Dim")

p_vae_sens <- ggplot(vae_comp_melt, aes(x = factor(Latent_Dim), y = value, 
                                        fill = variable)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "VAE Performance: Alternative Latent Dimensions",
    x = "Latent Dimension",
    y = "Final Loss",
    fill = "Loss Type"
  ) +
  scale_fill_manual(values = c("steelblue", "darkorange"),
                   labels = c("Training", "Validation")) +
  theme_bw(base_size = 14) +
  theme(plot.title = element_text(face = "bold"))

ggsave("figures/sensitivity_vae_architectures.png", p_vae_sens, 
       width = 10, height = 6, dpi = 300)
```

```{r}
# ---- Alternative Clustering Solutions (k-1, k+1) ----
cat("Testing alternative number of clusters...\n")

k_alternatives <- c(optimal_k - 1, optimal_k, optimal_k + 1)
k_comparison <- data.frame(
  K = integer(),
  Mean_Silhouette = numeric(),
  Total_WSS = numeric()
)

for (k_test in k_alternatives) {
  if (k_test < 2) next  # Skip if k < 2
  
  set.seed(123)
  km_test <- kmeans(df_clust_scaled, centers = k_test, nstart = 50, iter.max = 300)
  sil_test <- silhouette(km_test$cluster, dist(df_clust_scaled))
  
  k_comparison <- rbind(k_comparison, data.frame(
    K = k_test,
    Mean_Silhouette = mean(sil_test[, 3]),
    Total_WSS = km_test$tot.withinss
  ))
}

cat("\n  Clustering Solutions Comparison:\n")
print(k_comparison)
cat("\n")

write.csv(k_comparison, "tables/sensitivity_k_selection.csv", row.names = FALSE)
```


```{r}
# ---- Robustness to Missing Data Pattern ----
cat("Assessing robustness to missing data...\n")

missing_pattern <- final %>%
  mutate(
    Has_NfL = !is.na(NfL),
    Has_GFAP = !is.na(GFAP),
    Has_AB = !is.na(AB42_40_ratio),
    Has_pTau = !is.na(pTau181_recode),
    Complete_Biomarkers = Has_NfL & Has_GFAP & Has_AB & Has_pTau
  )

missing_summary <- missing_pattern %>%
  summarise(
    N_Total = n(),
    N_Complete = sum(Complete_Biomarkers),
    Pct_Complete = mean(Complete_Biomarkers) * 100,
    
    N_Missing_NfL = sum(!Has_NfL),
    N_Missing_GFAP = sum(!Has_GFAP),
    N_Missing_AB = sum(!Has_AB),
    N_Missing_pTau = sum(!Has_pTau)
  )

cat("\n  Missing Data Summary:\n")
print(missing_summary)
cat("\n")

write.csv(missing_summary, "tables/sensitivity_missing_data.csv", row.names = FALSE)

# Check if missingness is related to demographics/outcomes
missing_pattern_test <- missing_pattern %>%
  group_by(Complete_Biomarkers) %>%
  summarise(
    N = n(),
    Mean_Age = mean(PAGE, na.rm = TRUE),
    Pct_Female = mean(GENDER == "Female", na.rm = TRUE) * 100,
    Mean_Cog = mean(Dementia_Score_Imputed_2016, na.rm = TRUE)
  )

write.csv(missing_pattern_test, "tables/sensitivity_missing_by_demographics.csv", 
          row.names = FALSE)
```


```{r}
# ---- Summary of All Sensitivity Analyses ----
sensitivity_summary <- data.frame(
  Analysis = c(
    "Alternative ATN cutoffs (quartile)",
    "VAE latent dimension selection",
    "Alternative clustering (k±1)",
    "Missing data pattern"
  ),
  Finding = c(
    paste0("ARI = ", round(ari_atn, 3), " with quartile cutoffs"),
    paste0("Optimal latent_dim = ", vae_comparison$Latent_Dim[which.min(vae_comparison$Final_Val_Loss)]),
    paste0("Best k = ", k_comparison$K[which.max(k_comparison$Mean_Silhouette)]),
    paste0(round(missing_summary$Pct_Complete, 1), "% with complete biomarkers")
  ),
  Interpretation = c(
    ifelse(ari_atn > 0.6, "High agreement", "Moderate agreement"),
    "Supports dimensionality choice",
    ifelse(k_comparison$K[which.max(k_comparison$Mean_Silhouette)] == optimal_k, 
           "Confirms k selection", "Suggests alternative k"),
    "Missing not at random - potential bias"
  )
)

knitr::kable(sensitivity_summary, 
             caption = "Summary of Sensitivity Analyses")

write.csv(sensitivity_summary, "tables/sensitivity_analyses_summary.csv", 
          row.names = FALSE)

cat("\n=== SENSITIVITY ANALYSES COMPLETE ===\n\n")
```


### Method Comparison Summary Table

```{r}
# ---- Comprehensive Method Comparison ----

cat("\n=== CREATING COMPREHENSIVE METHOD COMPARISON ===\n")

# Gather all metrics
method_comparison <- data.frame(
  Method = c("ATN Framework", "K-Means Clustering", "VAE Latent Space", "PCA"),
  
  Approach = c(
    "Theory-driven",
    "Data-driven (unsupervised)",
    "Data-driven (deep learning)",
    "Data-driven (linear)"
  ),
  
  N_Groups = c(
    length(unique(final$ATN)),
    optimal_k,
    NA,
    NA
  ),
  
  Mean_Silhouette = c(
    NA,
    mean_sil_original,
    mean_sil_vae,
    mean_sil_pca
  ),
  
  Variance_Explained = c(
    NA,
    NA,
    NA,
    round(sum(var_pca) * 100, 1)
  ),
  
  Cognitive_Association = c(
    format_p(kw_cog$p.value),
    format_p(kruskal.test(Dementia_Score_Imputed_2016 ~ Cluster, 
                         data = final_tab)$p.value),
    NA,
    NA
  ),
  
  Interpretability = c("High", "Medium", "Low", "High"),
  
  Advantages = c(
    "Clinical relevance; validated cutoffs",
    "Discovers natural groupings; no assumptions",
    "Non-linear relationships; dimensionality reduction",
    "Simple; variance maximization"
  ),
  
  Limitations = c(
    "Binary cutoffs lose information; platform-specific",
    "Requires k selection; assumes spherical clusters",
    "Black box; requires tuning",
    "Linear only; may miss non-linear patterns"
  )
)

knitr::kable(method_comparison, 
             caption = "Comprehensive Comparison of Classification Approaches")

write.csv(method_comparison, "tables/method_comparison_comprehensive.csv", 
          row.names = FALSE)
```


```{r}
# Agreement/concordance matrix
agreement_matrix <- data.frame(
  Comparison = c(
    "ATN vs K-Means",
    "ATN vs ATN_quartile",
    "K-Means (4D) vs PCA (2D)",
    "K-Means (4D) vs VAE (2D)"
  ),
  ARI = c(
    ari_val,
    ari_atn,
    adjustedRandIndex(
      km$cluster,
      kmeans(pca_full$x[, 1:2], centers = optimal_k, nstart = 50)$cluster
    ),
    adjustedRandIndex(
      df_clust_clean$Cluster[match(latent_vae_complete$HHID_PN, 
                                    df_clust_clean$HHID_PN)],
      kmeans(as.matrix(latent_vae_complete[, c("z1", "z2")]), 
             centers = optimal_k, nstart = 50)$cluster
    )
  ),
  NMI = c(
    nmi_val,
    nmi_atn,
    NA,
    NA
  ),
  Interpretation = c(
    ifelse(ari_val > 0.3, "Moderate agreement", "Low agreement"),
    ifelse(ari_atn > 0.6, "High agreement", "Moderate agreement"),
    "Compares projection methods",
    "Compares projection methods"
  )
)

knitr::kable(agreement_matrix, digits = 3,
             caption = "Agreement Between Classification Methods")

write.csv(agreement_matrix, "tables/method_agreement_matrix.csv", row.names = FALSE)

cat("\n=== METHOD COMPARISON COMPLETE ===\n\n")
```


### Final Summary Report

```{r}
# ---- Generate Final Summary Report ----

cat("\n========================================\n")
cat("FINAL ANALYSIS SUMMARY\n")
cat("========================================\n\n")

# Sample characteristics
cat("1. SAMPLE CHARACTERISTICS\n")
cat("   Total participants:", nrow(final), "\n")
cat("   With complete biomarkers:", nrow(df_clust), "\n")
cat("   With complete data (biomarkers + covariates):", nrow(final_tab), "\n\n")

# ATN distribution
cat("2. ATN FRAMEWORK\n")
cat("   Number of ATN profiles:", length(unique(final$ATN)), "\n")
cat("   Most common profile:", names(which.max(table(final$ATN))), 
    "(n =", max(table(final$ATN)), ")\n")
cat("   Least common profile:", names(which.min(table(final$ATN))), 
    "(n =", min(table(final$ATN)), ")\n\n")

# Clustering results
cat("3. DATA-DRIVEN CLUSTERING\n")
cat("   Optimal k:", optimal_k, "\n")
cat("   Mean silhouette width:", round(mean_sil_original, 3), "\n")
cat("   Stability (mean bootstrap ARI):", round(mean(ari_scores), 3), "\n\n")

# Agreement
cat("4. ATN-CLUSTER AGREEMENT\n")
cat("   Adjusted Rand Index:", round(ari_val, 3), "\n")
cat("   Normalized Mutual Information:", round(nmi_val, 3), "\n")
cat("   Interpretation:", 
    ifelse(ari_val > 0.3, "Moderate agreement", "Low agreement"), "\n\n")

# Predictive performance
cat("5. BIOMARKER PREDICTIVE PERFORMANCE (Test Set)\n")
for (i in 1:nrow(test_performance)) {
  cat("   ", test_performance$Biomarker[i], "AUC:", 
      test_performance$Test_AUC[i], "\n")
}
cat("\n")

# Dimensionality reduction
cat("6. DIMENSIONALITY REDUCTION\n")
cat("   PCA (2D) variance explained:", round(sum(var_pca) * 100, 1), "%\n")
cat("   VAE mean silhouette:", round(mean_sil_vae, 3), "\n")
cat("   PCA mean silhouette:", round(mean_sil_pca, 3), "\n\n")

# Key findings
cat("7. KEY FINDINGS\n")
cat("   • ATN and data-driven clusters show", 
    ifelse(ari_val > 0.3, "moderate", "limited"), "alignment\n")
cat("   • Best performing biomarker:", 
    test_performance$Biomarker[which.max(test_performance$Test_AUC)], 
    "(AUC =", max(test_performance$Test_AUC), ")\n")
cat("   • Clustering is", 
    ifelse(mean(ari_scores) > 0.7, "highly stable", 
           ifelse(mean(ari_scores) > 0.5, "moderately stable", "unstable")),
    "\n")
cat("   • VAE", 
    ifelse(mean_sil_vae > mean_sil_pca, "outperforms", "underperforms"),
    "PCA for cluster separation\n\n")

cat("8. OUTPUT FILES GENERATED\n")
cat("   Tables:", length(list.files("tables", pattern = "\\.csv$")), "files\n")
cat("   Figures:", length(list.files("figures", pattern = "\\.png$")), "files\n")
cat("   Results:", length(list.files("results")), "files\n\n")

cat("========================================\n")
cat("ANALYSIS COMPLETE\n")
cat("========================================\n\n")
```


```{r}
# Save summary report
summary_report <- list(
  sample_size = nrow(final),
  complete_biomarkers = nrow(df_clust),
  optimal_k = optimal_k,
  mean_silhouette = mean_sil_original,
  ari_atn_cluster = ari_val,
  nmi_atn_cluster = nmi_val,
  best_biomarker = test_performance$Biomarker[which.max(test_performance$Test_AUC)],
  best_auc = max(test_performance$Test_AUC),
  clustering_stability = mean(ari_scores),
  pca_variance = sum(var_pca),
  vae_silhouette = mean_sil_vae,
  pca_silhouette = mean_sil_pca
)

saveRDS(summary_report, "results/analysis_summary.rds")

# Create a summary table
final_summary_table <- data.frame(
  Metric = names(summary_report),
  Value = unlist(summary_report)
)

write.csv(final_summary_table, "tables/analysis_summary.csv", row.names = FALSE)
```


```{r}
# ---- Transcriptomics Integration (I did not work on this part of the project) ----
# transcriptome_data <- read_csv("TheDataSet.csv") # I never had access to this data the team that did it had access to it

# final_omics <- final %>%
#   inner_join(transcriptome_data, by = "HHID_PN")

# gene_cols <- grep("^gene", colnames(final_omics), value = TRUE)

# dds <- DESeqDataSetFromMatrix(
#   countData = as.matrix(final_omics[, gene_cols]),
#   colData = final_omics[, c("ATN")],
#   design = ~ ATN
# )

# dds <- DESeq(dds)
# res <- results(dds)
# sig <- res[which(res$padj < 0.05), ]
# head(sig)
```
